Automated topic naming to support analysis of software maintenance activities

### 2010-08-20 icse/icse2011.tex
write up for analysis of timelines
Signed-off-by: Neil Ernst <neil@neilernst.net>

\begin{comment}
A quick mind dump before I forget:
* portability was very popular
* there were many instances of concurrency
* There were often topics of crap commits with no comment
* might need more than the top words to compare
* Things like tests were noticable more by file type
* Configuration management is not part of the software quality ontology?
* I didn't tag none I tagged unknown, so that's some inconsistency between us
* There were a lot of copyright changes in mysql
* mysql had optimizer module which did not mean optimization
* I tried to assign reliability/bug for reliability bugs
  and Functionality/Correctness for fixes
* Some security changes came later
* Mysql 3.23 transitioned from a project being developed to
  a project being maintained, more frequent back-port bug-fixes
\end{comment}

\subsection{Summary of techniques}
While unsupervised techniques (LSI and LDA are both unsupervised) are appealing in their lack of human intervention, and thus lower effort, supervised learners have the advantage of domain knowledge (training sets) which typically means improved results. 
Our manual annotations were fairly quick to do, taking on average five minutes per 30 day period.

@@

We found that the multi-label learners of BR, CLR and HOMER did not do as well for Macro-ROC as NaiveBayes and other naive Bayes derived learners. 
This suggests that by combining together multiple NaiveBayes learners we could probably label sets of topics effectively, but it would require a separate NaiveBayes learner per label.

\section{Using \textbf{Nomen} to understand software maintenance activities} 
\label{sec:analysis}
% \todo[inline]{talk about this topic in one section}
% \todo[inline]{reference project name Nomen more}

Using the classification results from the previous sections, we evaluated two research questions:
\begin{enumerate}
\item Do label frequencies change over time? That is, is a certain quality of more interest at one point in the life-cycle than some other? 
\item  Do the different projects differ in their relative topic interest? That is, is a particular quality more important to one project than the other projects?  
\end{enumerate}

\subsection{Visualizations}
Figures \ref{fig:maxdb-timeline} and \ref{fig:mysql-timeline} show the temporal patterns of label frequencies. 
There are two measures represented. 
One, the relative frequency, shown in the grey boxes, represents the number of labels with that NFR in that period, 
relative to the maximum number of topics assigned to that NFR. 
The second, absolute frequency, compares the number of topics labelled with that NFR per period 
relative to the maximum number of labelled topics overall, for that project.

\textbf{Understanding the figures} -- For example, in Figure \ref{fig:mysql-timeline}, we see that the NFRs \emph{functionality}, \emph{portability} and \emph{maintainability} contain more labeled topics, since these NFRs have been more intensely shaded. % boxes. not sure boxes is important..
%We note that the gap in the MaxDB data is due to an interruption in the source data, having to do with the infrastructure for releasing developer artifacts, rather than specific occurrences in the project.
% I grepped query1.mrs2 http://softwareprocess.es/y/y/maxdb-76.tar.gz
% and                   http://softwareprocess.es/y/maxdb-7500.tar.gz
The gap in the MaxDB is due to the fact that other versions of MaxDB exist on different repositories and during this period from February 2005 to Jun 2005 the focus of development shifted to MaxDB 7.6. 
Immediately after this period of inactivity, there was a large deluge of commits added to the MaxDB 7.500 repositories.
%\todo[inline]{can I pin this down more ?}
There are more \textsf{None} labels in MaxDB because a number of topics were to do with code cleanup or automated checking using tools like \texttt{sutcheck}, a tool-specific to the development process at MaxDB.

%XXX 
\textbf{Timeline of key events} -- Figs. \ref{} show key events for each of our projects.  % where's the figure?
Several external events (external to `normal' development activity, that is) affected our projects. 
For example, both projects show a high number of NFRs recognized at the first period of analysis. 
This is due to our window choice: we deliberately targeted our analysis to 
when MySQL 3.23 was first announced %Major update \& End-of life
and when MaxDB 7.500 was first announced.

Note that MySQL's release process was haphazard at this point: they would release a patch version whenever the need arose. 
By version 5.1, the company moved to a milestone release, with new releases targeted for every six months.

\todo[inline]{Does our visualization capture the entire timeline?}

%\todo[inline]{try to pin down these reasons below}
%  MySQL 3.23 was put on maintenance.
% 
% MySQL forks repositories per every version and 3.23 was supported for a
% long time. there was 4.0 4.1 5.0 5.1
% 
% This is the MySQL 3.23 branch
% 
% Same thing with MaxDB 7.5
% 7.6 is in a different repository

% For MaxDB (Figure \ref{fig:maxdb}) the largest trends were related to maintainability and portability. MaxDB supports numerous platforms and portability was a constant issue facing the project's development. MySQL was different, with many topics overlapping, so there were many different subsets. Functionality trends were prevalent throughout its history (the largest reddish streak across the top). Two combination  tags of ``functionality portability'' (orange) and ``maintainability portability'' (teal) streaked across most of the history of MySQL. Since this was MySQL 3.23, a stable branch, we expect that issues dealing with portability and maintenance would be the primary concerns of this branch: developers probably wanted it to work with current systems and thus had to update it.

%\todo[inline]{Update this}

\begin{figure*}[ht]
\centering
\subfloat[MySQL 3.23]{
\includegraphics[width=\textwidth]{figures/mysql-timeline}
\label{fig:maxdb-timeline}                               
}

\subfloat[MaxDB 7.500]{                                 
\includegraphics[width=\textwidth]{figures/maxdb-timeline}
\label{fig:mysql-timeline}
}
\caption[]{NFR label per period. Each box represents a 30-day period. Numbers represent number of topics in that period labelled with that NFR. Box intensity (saturation) is mapped to label frequency relative to the largest label value of \emph{all} NFRs. Grey bars indicate label frequency relative to \emph{a particular} NFR.
% \todo[inline]{better discussion}
}
\label{fig:timelines}
\end{figure*}

% 
% \begin{figure*}
%   \centering
%   % hi-res version
%   %\includegraphics[height=.55\textheight]{maxdb-time-smear-plot-Equal}
%   \includegraphics[width=\textwidth]{figures/maxdb-timeline-all}
%   % \includegraphics[height=.57\textheight]{small-maxdb-time-smear-plot-Equal}
%  % \caption{MaxDB 7.500: Labelled topics related to non-functional software qualities plotted over time, continuous topics are trends that occur across time}
%   \caption{MaxDB 7.500: NFRs over time. Numbers represent topics labelled with that quality. Box intensity (saturation) is mapped to label frequency relative to the largest label value.}
%  \label{fig:maxdb}
% \end{figure*}
% 
% \begin{figure*}%[h]
%   \centering
%  %\includegraphics[width=.97\textwidth]{mysql-time-smear-plot-Equal}
%   %\includegraphics[width=.97\textwidth]{small-mysql-time-smear-plot-Equal}
%   \includegraphics[width=\textwidth]{figures/mysql-timeline-all}
%   \caption{MySQL 3.23: NFRs over time. Topic frequency relative to that quality is mapped to alpha (transparency) value.}
%   \label{fig:mysql}
% \end{figure*}

>>>>

\subsection{Summary of techniques}
While unsupervised techniques such as LDA is appealing in its lack of human intervention, and thus lower effort, supervised learners have the advantage of domain knowledge which typically means improved results. Our manual annotations were fairly quick to do, taking on average five minutes per 30 day period.

@@

We found that the multi-label learners of BR, CLR and HOMER did not do as well for Macro-ROC as NaiveBayes and other NaiveBayes derived learners. 
This suggests that by sticking together multiple NaiveBayes learners we could probably label sets of topics effectively, but it would require a separate NaiveBayes learner per label.

\section{Using \textbf{Nomen} to understand software maintenance activities} 
\label{sec:analysis}
% \todo[inline]{talk about this topic in one section}
% \todo[inline]{reference project name Nomen more}
Since our ROC values produced acceptable results, we now turn to \emph{applying} those techniques to understand software maintenance activities. We evaluated two research questions:
\begin{enumerate}
\item Do label frequencies change over time? That is, is a certain quality of more interest at one point in the life-cycle than some other? 
\item  Do the different projects differ in their relative topic interest? That is, is a particular quality more important to one project than the other projects?  
\end{enumerate}

There are several possible applications of \textsf{Nomen}'s techniques. 
For outside users of the software, \textsf{Nomen} can provide insight into that company's development practices. 
Are they focused on usability, for example?
Within the organization \textsf{Nomen} acts as a historical sanity check. 
Confirmation bias is the difference between what people think happened and what actually happened. 
Some software releases, for instance, are poorly received since they are quite buggy. 
\textsf{Nomen} would help explain why this was the case, since it can identify NFRs which are dominating developer attention.

\subsection{Timeline visualization}
Figures \ref{fig:maxdb-timeline} and \ref{fig:mysql-timeline} show the temporal patterns of label frequencies. 
There are two measures represented. 
One, the relative frequency, shown in the grey boxes, represents the number of labels with that NFR in that period, 
relative to the maximum number of topics assigned to that NFR. 
The second, absolute frequency, compares the number of topics labelled with that NFR per period 
relative to the maximum number of labelled topics overall, for that project. The topmost row in the diagrams is reserved for historical events of note. We refer to these occurrences in the discussion which follows.

\textbf{Understanding the figures} -- In Fig. \ref{fig:mysql-timeline}, we see that the NFRs \emph{functionality}, \emph{portability} and \emph{maintainability} contain more labeled topics, since these NFRs have been more intensely shaded. % boxes. not sure boxes is important..
%We note that the gap in the MaxDB data is due to an interruption in the source data, having to do with the infrastructure for releasing developer artifacts, rather than specific occurrences in the project.
% I grepped query1.mrs2 http://softwareprocess.es/y/y/maxdb-76.tar.gz
% and                   http://softwareprocess.es/y/maxdb-7500.tar.gz
There are more \textsf{None} labels in MaxDB because a number of topics were to do with code cleanup or automated checking using tools like \texttt{sutcheck}, a tool-specific to the development process at MaxDB.

\textbf{Timeline of key events} -- As discussed earlier, the top row in each figure shows key events for each project. \textsf{Nomen} can pick out the underlying NFR activity behind these events. For example, both projects show a high number of NFRs recognized at the first period of analysis. 
This is due to our window choice: we deliberately targeted our analysis to coincide with the development lifecycle of a particular major release of the software. For MaxDB, version 7.5.00  was released in December of 2003. 
From analyzing the mailing list for external validation, we know that release 7.5.00.23 saw the development of PHP interfaces, possibly accounting for the simultaneous increase in the \emph{portability} NFR at the same time.
The gap in MaxDB (Fig. \ref{fig:maxdb-timeline}) is due to a shift in development focus (from February 2005 to Jun 2005) to MaxDB 7.6, which is released in June 2005.

For MySQL (Fig. \ref{fig:mysql-timeline}), we similarly validated our NFR patterns with external mailing list data. This release of MySQL was the first to be licenced under the GPL. Version 3.23.31 (January, 2001) was the production release (non-beta), and we see a flurry of topics labeled with \emph{functionality} and \emph{maintainability}, as one might expect. After this point, we can consider this version to enter the maintenance phase of the lifecycle.

%Note that MySQL's release process was haphazard at this point: they would release a patch version whenever the need arose. By version 5.1, the company moved to a milestone release, with new releases targeted for every six months.

% \todo[inline]{Does our visualization capture the entire timeline?}

%\todo[inline]{try to pin down these reasons below}

\begin{figure*}[ht]
\centering
\subfloat[MySQL 3.23]{
\includegraphics[width=\textwidth]{figures/mysql-timeline}
\label{fig:mysql-timeline}                               
}

\subfloat[MaxDB 7.500]{                                 
\includegraphics[width=\textwidth]{figures/maxdb-timeline}
\label{fig:maxdb-timeline}
}
\caption[]{NFR label per period. Each box represents a 30-day period. Numbers represent number of topics in that period labelled with that NFR. Box intensity (saturation) is mapped to label frequency relative to the largest label value of \emph{all} NFRs. Grey bars indicate label frequency relative to \emph{a particular} NFR.
% \todo[inline]{better discussion}
}
\label{fig:timelines}
\end{figure*}
